---
title: "Daily Exercise 05"
author: "Mike Johnson"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## COVID-19 Data

### Data

We are going to practice some data wrangling skills using a real-world dataset about COVID cases curated and maintained by the New York Times. The data has been used to create reports and data visualizations like [this](https://www.nytimes.com/interactive/2020/us/coronavirus-spread.html?referringSource=articleShare), and are archived on a GitHub repo [here](https://github.com/nytimes/covid-19-data). Looking at the README in this repository we read:

> "We are providing two sets of data with cumulative counts of coronavirus cases and deaths: one with our most current numbers for each geography and another with historical data showing the tally for each day for each geography ... the historical files are the final counts at the end of each day ... The historical and live data are released in three files, one for each of these geographic levels: U.S., states and counties. A smaller file with only the most recent 30 days of data is also available"

For this lab we will use the historic, recent, country level data which is stored as an updating CSV at this URL:


```{r, eval = FALSE}
https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties-recent.csv
```

### Start

I will get you started this week, in the following code chunk, I am attaching the tidyverse package; saving the NY-Times URL as a value called "url"; and I am reading that URL into an object called `covid`

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
url = 'https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties-recent.csv'
covid = read_csv(url)
head(covid, 5)
```

Hint: You can print the top X rows of a `data.frame` with `slice`.

```{r, eval = FALSE}
slice(covid, 1:5)
```

to print the top 5 columns of the raw covid `object` 

## Question 1: 

Use `dplyr` verbs to  create a data.frame of the **5** counties with the most current cases. Remember, the cases and deaths are cumulative, so you only need to deal with the data for the most current (`max`) date.

(**Hint:** `filter`, `arrange`,  `slice`)

```{r}
most_cases = covid  #assign to new dataframe that I can freely edit for this question
most_cases %>%   #create pipeline
  filter(cases > 0) %>% #eliminate nulls
  filter(date == max(date)) %>% #only the latest date with cumulative cases
  filter(county != "Unknown") %>% #filter out unknown counties
  arrange(-cases) %>% #sort by cases in decreasing order
  slice(1:5) #pick top 5
```

## Question 2: 

Use `dplyr` verbs to create a data.frame of the **5** states with the _most_ cases current cases. 

(**Hint:** `filter`, `group_by`, `summarize`, `arrange`, `slice`)

```{r}
state_cases = covid %>%   #new datafram for new question & start pipeline
  filter(cases > 0) %>%  #eliminate nulls
  filter(date == max(date)) %>% #pick only the final date with cumulative cases
  group_by(state) %>% #group into state
  summarize(total_cases = sum(cases, na.rm= FALSE)) %>% # add up all the cases by state into a new table
  arrange(-total_cases) %>%  #sort in decreasing order
  slice(1:5) #choose top 5

(state_cases)
```
This correlates with the current 33.6 million total US cases from Google. I had to double check since I made the error of not filtering by max date and ended up with 1 billion cases total and around 300 million in California alone.


## Question 3: 

Use the `dplyr` verbs to report the **5** counties with the _worst_ current death/cases ratio: (e.g.$100* (deaths/cases)$)

(**Hint:** You will need to remove those where `cases == 0` and `county == Unknown`)
(**Hint:** `filter`, `mutate`, `arrange`, `slice`)

```{r}
worst_county_deaths = covid %>%  #new dataframe for new question & start pipeline
  filter(deaths > 0 & cases > 0) %>% #eliminate nulls & zeros
  filter(date == max(date)) %>% #select cumulative data
  filter(county != "Unknown") %>% #filter out unknown counties
  group_by(county) %>% #group counties
  summarize(percent_death_rate = 100*(sum(deaths, na.rm = FALSE) / sum(cases, na.rm = FALSE))) %>% #death rate by county
  arrange(-percent_death_rate) %>% #decending death rate
  slice(1:5)
(worst_county_deaths)
```

## Question 4: 

Use the `dplyr` verbs to report the **5** states with the _worst_ current death/case ratio.

(**Hint:** `filter`, `group_by`, `summarize`, `mutate`, `arrange`, `slice`)

```{r}
worst_state_deaths = covid %>%  #new datafream and pipeline
  filter(deaths > 0 & cases > 0) %>% #eliminate nulls & zeros
  filter(date == max(date)) %>% #select cumulative data
  filter(county != "Unknown") %>% #filter out unknown counties
  group_by(state) %>% #group state
  summarize(percent_death_rate = 100*(sum(deaths, na.rm = FALSE) / sum(cases, na.rm = FALSE))) %>% #death rate by state
  arrange(-percent_death_rate) %>% #decending death rate
  slice(1:5)
(worst_state_deaths)

```
Something is weird with Puerto Rico, I'm going to investigate.

```{r}
PR_covid = covid
PR_covid %>% 
  filter(state == "Puerto Rico") %>% #I only want PR
  filter(date == max(date)) %>% #only cumulative
  filter(deaths > 0 & cases > 0) #eliminate nulls & zeros
```
Okay, so this explains a lot, mostly about an error in data gathering. In this case I should go back and filter out all counties and states that have unknown in them now that I know that this is the convention of the data set for NA names.
# END



